Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-12-15 paper/manuscript.tex
Mention and cite Macuglia review

After an alchemical calculation is performed, which generally involves multiple simulations at a variety of alchemical states, the data must be analyzed to compute an estimate of the free energy for the transformation of interest.
 Early work used simple but statistically suboptimal estimators for this: free energy perturbation (FEP) used a simple (but highly biased) estimator based on the Zwanzig relation~\cite{zwanzig1954hightemperature} or numerical quadrature via thermodynamic integration (TI), for which the theory dates back the better part of a century but with the first computational applications emerging in the 1980's and 90's~\cite{kirkwood1935statistical, jorgensen1985monte, kollman1993free, wong1986dynamics, merz1989free}. %
 More recent developments have seen new, highly efficient statistical estimators that make better use of all the data, often building on the more efficient and less biased Bennett acceptance ratio (BAR)~\cite{bennett1976efficient}, producing multistate generalizations~\cite{shirts2008statisticallya} or removing the need for global equilibrium~\cite{wu2016multiensemble, mey2014xtram, wu2014statistically}.

@@

\subsubsection{Relative binding free energy (RBFE)}

@@

Alchemical free energy calculations involving small molecules seem to achieve, in favorable cases, root mean square (RMS) errors relative to experiment around 1-2 kcal/mol depending on force field, system, and a variety of other factors such as simulation time, sampling method, and whether the calculations employed are absolute or relative. A small selection of example datasets and case studies can be found in Sec.~\ref{sec:benchmark} at the end of this document.
However, the domain of applicability is a significant concern~\cite{sherborne2016collaborating, cournia2017relative}, especially for relative calculations, which typically require a high quality and usually experimental bound structure of a closely
related ligand as a starting point. Additional factors such as slow protein or ligand rearrangements, uncertainties in ligand binding mode, or charged ligands can make these calculations far less reliable and more of a research effort.

@@

\paragraph{Topologies} In regular biomolecular simulations one studies physical systems or, in other words, all entities in the system, e.g., protein, ligand and waters, represent real molecules. For the following, it is helpful to classify such simulation systems as having a {\em chemical} topology. In relative free energy calculations, however, one has to transform, e.g., ligand A into ligand B etc. (see \sbnote{appropriate cross-reference!}). The required alchemical transformations can be realized in several ways, each with distinct advantages and disadvantages. In the first relative free energy calculations, these transformations were achieved with the help of chimeric entities, capable of representing both ligand A or ligand B. The important point here is that there is a single set of coordinates for this chimeric ligand, but its ``topology'' is flexible and can, for the same set of coordinates, represent either end state. We will consider this class of approaches as using an {\em alchemical} topology with a single set of coordinates. The construction of these chimeric intermediates quickly gets complicated and is difficult to automate; therefore, approaches have been developed in which both end states (e.g., ligand A and B) are present simultaneously. Thus, we have dual coordinates, but chemical topologies for each of the entities. In the following, we outline these approaches in more detail, together with additional considerations how the various approaches differ in practical use. While we will point out the respective advantages and disadavantages, we stress that the user's option in practice often are limited by the choice of simulation software.

>>>>

After an alchemical calculation is performed, which generally involves multiple simulations at a variety of alchemical states, the data must be analyzed to compute an estimate of the free energy for the transformation of interest.
 Early work used simple but statistically suboptimal estimators for this: free energy perturbation (FEP) used a simple (but highly biased) estimator based on the Zwanzig relation~\cite{zwanzig1954hightemperature} or numerical quadrature via thermodynamic integration (TI), for which the theory dates back the better part of a century but with the first computational applications emerging in the 1980's and 90's~\cite{kirkwood1935statistical, jorgensen1985monte, kollman1993free, wong1986dynamics, merz1989free}. Macuglia et al. review the evolution of biomolecular alchemical free energy calculations into practically useful tools \cite{macuglia2025dawn}. %
 More recent developments have seen new, highly efficient statistical estimators that make better use of all the data, often building on the more efficient and less biased Bennett acceptance ratio (BAR)~\cite{bennett1976efficient}, producing multistate generalizations~\cite{shirts2008statisticallya} or removing the need for global equilibrium~\cite{wu2016multiensemble, mey2014xtram, wu2014statistically}.

@@

\subsubsection{Relative binding free energy (RBFE)}
\label{sec:theory-rbfe}

@@

Alchemical free energy calculations involving small molecules seem to achieve, in favorable cases, root mean square (RMS) errors relative to experiment around 1-2 kcal/mol depending on force field, system, and a variety of other factors such as simulation time, sampling method, and whether the calculations employed are absolute or relative. Example datasets and case studies are available in the companion article on best practices for large-scale alchemical free energy calculations.
However, the domain of applicability is a significant concern~\cite{sherborne2016collaborating, cournia2017relative}, especially for relative calculations, which typically require a high quality and usually experimental bound structure of a closely
related ligand as a starting point. Additional factors such as slow protein or ligand rearrangements, uncertainties in ligand binding mode, or charged ligands can make these calculations far less reliable and more of a research effort.

@@

\paragraph{Topologies} In regular biomolecular simulations one studies physical systems or, in other words, all entities in the system, e.g., protein, ligand and waters, represent real molecules. For the following, it is helpful to classify such simulation systems as having a {\em chemical} topology. In relative free energy calculations, however, one has to transform, e.g., ligand A into ligand B etc. (see Sec.~\ref{sec:theory-partitioning}). The required alchemical transformations can be realized in several ways, each with distinct advantages and disadvantages. In the first relative free energy calculations, these transformations were achieved with the help of chimeric entities, capable of representing both ligand A or ligand B. The important point here is that there is a single set of coordinates for this chimeric ligand, but its ``topology'' is flexible and can, for the same set of coordinates, represent either end state. We will consider this class of approaches as using an {\em alchemical} topology with a single set of coordinates. The construction of these chimeric intermediates quickly gets complicated and is difficult to automate; therefore, approaches have been developed in which both end states (e.g., ligand A and B) are present simultaneously. Thus, we have dual coordinates, but chemical topologies for each of the entities. In the following, we outline these approaches in more detail, together with additional considerations how the various approaches differ in practical use. While we will point out the respective advantages and disadavantages, we stress that the user's option in practice often are limited by the choice of simulation software.
